---
title: "cgMLST distance matrix"
author: "Mona Taouk"
output: html_document
date: "2023-03-29"
---

# Libraries and load data

```{r, warning=F}
load("Threshold.RData")

# Libraries
library(ggtree)
library(tidyverse)
library(ggplot2)
library(RColorBrewer)
library(ggridges)
library(reshape2)
library(scales)
library(patchwork)
library(ape)
```

#### Set up matrix, melt, merge and add columns

```{r, warning=F}

# melt matrix keeping only top of pyramid
cgMLST_melt = data.frame(ID1=rownames(cgMLST_matrix_all)[row(cgMLST_matrix_all)[upper.tri(cgMLST_matrix_all)]], ID2=colnames(cgMLST_matrix_all)[col(cgMLST_matrix_all)[upper.tri(cgMLST_matrix_all)]], dist=cgMLST_matrix_all[upper.tri(cgMLST_matrix_all)])

# Merge cgMLST_melt with dataframe that tells us which IDs are the callibration isolates
cgMLST_melt = cgMLST_melt %>% merge(intrapatient, by.x = "ID1", by.y = "AUSMDU", all.x = F, all.y = F) %>% merge(intrapatient, by.x = "ID2", by.y = "AUSMDU", all.x = F, all.y = F)

# Check for NAs
sum(is.na(cgMLST_melt))

# Create a new column called "check" that indicates whether the Patient_no.x and Patient_no.y columns are the same or different
cgMLST_melt$check = ifelse(cgMLST_melt$Patient_no.x == cgMLST_melt$Patient_no.y, "same", "diff")

```

#### make dataframes for between, within etc

```{r, warning=F}

between_patients = filter(cgMLST_melt, cgMLST_melt$check == "diff")
between_patients$legend = "Between individuals"
between_patients = data.frame(between_patients$dist, between_patients$legend)
colnames(between_patients)[1] <- "dist"
colnames(between_patients)[2] <- "legend"
min(between_patients$dist)
max(between_patients$dist)
median(between_patients$dist)

within_patients = filter(cgMLST_melt, cgMLST_melt$check == "same")
within_patients = filter(within_patients, within_patients$Cat.x == "WIthin-patient")
within_patients$legend = "Within individuals"
within_patients = data.frame(within_patients$dist, within_patients$legend)
colnames(within_patients)[1] <- "dist"
colnames(within_patients)[2] <- "legend"
min(within_patients$dist)
max(within_patients$dist)
median(within_patients$dist)

within_site = filter(cgMLST_melt, cgMLST_melt$check == "same")
within_site = filter(within_site, within_site$Cat.x == "Within-site")
within_site$legend = "Within site"
within_site = data.frame(within_site$dist, within_site$legend)
colnames(within_site)[1] <- "dist"
colnames(within_site)[2] <- "legend"
min(within_site$dist)
max(within_site$dist)
median(within_site$dist)

dist = c(cgMLST_matrix_all["AUSMDU00054152","AUSMDU00054265"], 
         cgMLST_matrix_all["AUSMDU00056203","AUSMDU00056226"],
         cgMLST_matrix_all["AUSMDU00058007","AUSMDU00058008"],
         cgMLST_matrix_all["AUSMDU00058728","AUSMDU00058842"],
         cgMLST_matrix_all["AUSMDU00059133","AUSMDU00059135"],
         cgMLST_matrix_all["AUSMDU00059905","AUSMDU00059919"],
         cgMLST_matrix_all["AUSMDU00060670","AUSMDU00060683"],
         cgMLST_matrix_all["AUSMDU00061345","AUSMDU00061360"],
         cgMLST_matrix_all["AUSMDU00064875","AUSMDU00064881"])
legend = c("Paired couples")
Pairs = data.frame(dist, legend)
min(Pairs$dist)
max(Pairs$dist)
median(Pairs$dist)

data_merged = rbind(between_patients, within_patients, within_site, Pairs)
data_merged_within = rbind(within_patients, within_site, Pairs)

rm(between_patients, within_patients, within_site, Pairs, dist, legend)
```

# Intra-patient threshold

```{r, warning=F}
ggplot(data_merged, aes(x=dist, fill = factor(legend, level = c("Between individuals", "Within individuals", "Within site", "Paired couples")))) +
  geom_histogram(binwidth = 1) +
  theme_bw() +
  theme(axis.text.x = element_text(size = 8, angle = 45, hjust = 1),
        legend.position = "none",
        panel.background = element_rect(color = "black", size = 1),
        strip.text.y = element_text(size = 8),
        strip.background = element_rect(color="black", size = 1),
        legend.title = element_text(size=8)) +
  ylab("") +
  xlab("Pairwise allelic difference") +
  scale_fill_brewer(palette = "greens") +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE, big.mark = " ")) +
  ggtitle("") +
  facet_grid(scales = "free", rows = vars(factor(legend, level = c("Between individuals", "Within individuals", "Within site", "Paired couples"))))


ggplot(data_merged, aes(x=dist, fill = factor(legend, level = c("Between individuals", "Within individuals", "Within site", "Paired couples")))) +
  geom_histogram(binwidth = 1) +
  theme_bw() +
  theme(axis.text.x = element_text(size = 8, angle = 45, hjust = 1),
        legend.position = "none",
        panel.background = element_rect(color = "black", size = 1),
        strip.text.y = element_text(size = 8),
        strip.background = element_rect(color="black", size = 1),
        legend.title = element_text(size=8)) +
  ylab("") +
  xlab("Pairwise cgMLST distance") +
  scale_fill_brewer(palette = "greens") +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE, big.mark = " ")) +
  ggtitle("") +
  facet_grid(scales = "free", rows = vars(factor(legend, level = c("Between individuals", "Within individuals", "Within site", "Paired couples")))) +
  scale_x_continuous(limits = c(-1, 10))
```

# Dates and distance correlation

## LSD Tree using 95% gaps ML tree

```{r}
ggtree(tree, layout = 'rectangular', mrsd="2020-06-30", linewidth = 0.25) + 
  theme_tree2() + 
  xlab("Time") +
  theme(panel.grid.major = element_line(color="darkgrey", size=.2),
        panel.grid.minor = element_line(color="grey", size=.2, linetype = "dashed"),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank()) +
  scale_x_continuous(breaks= c(1880, 1900, 1920, 1940, 1960, 1980, 2000, 2020))
```

```{r}
# Pairwise branch length distances for each pair of isolates
dist_mat <- cophenetic(tree)
row_sorted <- sort(rownames(dist_mat))
col_sorted <- sort(colnames(dist_mat))
dist_mat_sorted <- dist_mat[row_sorted, col_sorted]
dist_mat_melt <- data.frame(ID1=rownames(dist_mat_sorted)[row(dist_mat_sorted)[upper.tri(dist_mat_sorted)]], ID2=colnames(dist_mat_sorted)[col(dist_mat_sorted)[upper.tri(dist_mat_sorted)]], dist=dist_mat_sorted[upper.tri(dist_mat_sorted)])
dist_mat_melt$ID <- paste(dist_mat_melt$ID1, dist_mat_melt$ID2, sep = "_")
dist_mat_melt <- dist_mat_melt[, -1]
dist_mat_melt <- dist_mat_melt[, -1]

# Pairwise cgMLST allelic different for each isolate
cgMLST_dist_sorted <- cgMLST_matrix_unique[row_sorted, col_sorted]
cgMLST_dist_melt <- data.frame(ID1=rownames(cgMLST_dist_sorted)[row(cgMLST_dist_sorted)[upper.tri(cgMLST_dist_sorted)]], ID2=colnames(cgMLST_dist_sorted)[col(cgMLST_dist_sorted)[upper.tri(cgMLST_dist_sorted)]], dist=cgMLST_dist_sorted[upper.tri(cgMLST_dist_sorted)])
cgMLST_dist_melt$ID <- paste(cgMLST_dist_melt$ID1, cgMLST_dist_melt$ID2, sep = "_")
cgMLST_dist_melt <- cgMLST_dist_melt[, -1]
cgMLST_dist_melt <- cgMLST_dist_melt[, -1]
colnames(cgMLST_dist_melt)[1] <- "cgMLST"

#Merge dataframes together 
branch_and_cgMLSTdist <- merge(cgMLST_dist_melt, dist_mat_melt, by = "ID", all.y = T, all.x = T)

# Remove temporary dataframes
rm(dist_mat, cgMLST_dist_sorted, dist_mat_sorted, dist_mat_melt, cgMLST_dist_sorted, col_sorted, row_sorted)


# Plot association
ggplot(branch_and_cgMLSTdist, aes(y = cgMLST, x = dist)) +
  geom_bin2d(bins = 1000) +
  geom_smooth(method = "lm", formula = y ~ x, color = "black", fill = "darkgrey", alpha = 0.8, se = TRUE) +
  theme_bw() +
  theme(axis.text.x = element_text(size = 8, angle = 45, hjust = 1),
        panel.background = element_rect(color = "black", size = 1),
        plot.title = element_text(size = 10),
        legend.key.size = unit(2.5, 'mm'),
        legend.text = element_text(size=8),
        legend.title = element_text(size=8)) +
  labs(x = "Branch length", y = "Pairwise allelic difference") +
  scale_fill_continuous(type = "viridis") +
  labs(fill = "Density") +
  ggtitle("")

```

```{r}
# Find the slope of the line 
model <- lm(cgMLST ~ dist, data = branch_and_cgMLSTdist)
coef(model)[2] / 52

# Calculate the 95% confidence interval for the slope coefficient
conf_interval <- confint(model, level = 0.95)
conf_interval  
coef(model)[2]
```

# Adjust distance for rate of cgMLST change

adjusted pairwise cgMLST distance = unadjusted pairwise cgMLST
distance - (cgMLST rate \* pairwise temporal distance)

```{r}
# melt the matrix of pairwise alleleic differences for just the 5,881 isolates that represent a unique infection.
cgMLST_melt_unique <- melt(cgMLST_matrix_unique, varnames = c("ID1", "ID2"), value.name = "dist")

#merge dates into the melted matrix (this may take ~10 minutes)
cgMLST_melt_unique = cgMLST_melt_unique %>% merge(dates, by.x = "ID1", by.y = "ID", all.x = F, all.y = F) %>% merge(dates, by.x = "ID2", by.y = "ID", all.x = F, all.y = F)
  
# Check for NAs
sum(is.na(cgMLST_melt_unique))

# Make dates dates
cgMLST_melt_unique$Date.x <- as.Date(cgMLST_melt_unique$Date.x, format = "%d/%m/%Y")
cgMLST_melt_unique$Date.y <- as.Date(cgMLST_melt_unique$Date.y, format = "%d/%m/%Y")

#calculate the pariwise temporal difference between isolates
cgMLST_melt_unique$date = abs(difftime(cgMLST_melt_unique$Date.x, cgMLST_melt_unique$Date.y, units = "weeks"))
cgMLST_melt_unique$weeks_apart <- as.numeric(cgMLST_melt_unique$date, units = "weeks")

# Adjust by coefficent
cgMLST_melt_unique$adjust = cgMLST_melt_unique$weeks_apart * 0.04994926  
cgMLST_melt_unique$dist_adjusted = cgMLST_melt_unique$dist - cgMLST_melt_unique$adjust

# Make all values positive
cgMLST_melt_unique$dist_adjusted_pos = ifelse(cgMLST_melt_unique$dist_adjusted < 0, "0", cgMLST_melt_unique$dist_adjusted)

# Make sure all values are numeric
cgMLST_melt_unique$dist_adjusted_pos = as.numeric(cgMLST_melt_unique$dist_adjusted_pos)

# Reshape data into matrix for next step
adjusted_matrix <- dcast(cgMLST_melt_unique, ID1 ~ ID2, value.var = "dist_adjusted_pos")
row.names(adjusted_matrix) <- adjusted_matrix$ID1
adjusted_matrix$ID1 <- NULL
adjusted_matrix = as.matrix(adjusted_matrix)
```

## Cluster using H-clust

```{r}
#without-threshold
unadjusted_dist <- as.dist(cgMLST_matrix_unique) 

unadjusted_hc <- hclust(unadjusted_dist, method = "single")

unadjusted_cut <- cutree(unadjusted_hc, h=7)

#convert to dataframe
unadjusted_cut_df <- as.data.frame(unadjusted_cut)

rm(unadjusted_cut)
rm(unadjusted_hc)
rm(unadjusted_dist)

#adjusted-threshold
adjusted_dist <- as.dist(adjusted_matrix)

adjusted_hc <- hclust(adjusted_dist, method = "single")

adjusted_cut <- cutree(adjusted_hc, h=7)

#convert to dataframe
adjusted_cut_df <- as.data.frame(adjusted_cut)

rm(adjusted_cut)
rm(adjusted_hc)
rm(adjusted_dist)

```
